Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-12 talk/pepm2011/paper.tex
Typos.  Swap $u$ and $v$ in \texttt{get}, to match the \texttt{set}.

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
one associated with $p_{6}$ would know that its \texttt{intval} field contains
the constant -100.

The following operations, that use $p_{5}$ and $p_{6}$ could then be
optimized using that knowledge:

@@

So far we have only described what happens when static objects are used in
operations that read and write their fields and in guards. When the static
object is used in any other operation, it cannot stay static. For example, when
a static object is stored in a globally accessible place, the object needs to
actually be allocated, as it might live longer than one iteration of the loop
and because the partial evaluator looses track of it. This means that the static
objects needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that actually allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizers produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well) This means that instead of the jump,
the following operations are emitted:

@@

The final optimized trace of the example can be seen in
Figure~\ref{fig:step1}.

@@

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\label{fig:semantics}
\caption{The Operational Semantics of Simplified Traces}
\end{figure*}

@@

Traces are lists of operations. The operations considered here are \texttt{new} (to make
a new object), \texttt{get} (to read a field out of an object), \texttt{set} (to write a field
into an object) and \texttt{guard\_class} (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are immediately initialized to a real
location, otherwise the trace is malformed.

@@

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment and a heap, which are potentially changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would
be malformed. The environment could not directly map variables to object,
because several variables can contain a pointer to the \emph{same} object. Thus
the "indirection" is needed to express sharing.

@@

The new operation creates a new object $(T,\mathrm{null},\mathrm{null})$, on the
heap under a fresh location $l$ and adds the result variable to the environment,
mapping to the new location $l$.

The \texttt{get} operation reads a field $F$ out of an object and adds the result
variable to the environment, mapping to the read location. The heap is
unchanged.

@@

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace to variables in the optimized trace (which are written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. Thus the environment of the
optimizer serves a function similar to that of the environment in the semantics.

The static heap is a partial function from $V^*$ into the
set of static objects, which are triples of a type, and two elements of $V^*$.
An variable $v^*$ is in the domain of the static heap $S$ as long as the
optimizer can fully keep track of the object. The image of $v^*$ is what is
statically known about the object stored in it, \ie its type and its fields. The
fields of objects in the static heap are also elements of $V^*$ (or null, for
short periods of time).

When the optimizer sees a new operation, it optimistically removes it and
assumes that the resulting object can stay static. The optimization for all
further operations is split into two cases. One case is for when the
involved variables are in the static heap, which means that the operation can be
performed at optimization time and removed from the trace. These rules mirror
the execution semantics closely. The other case is that nothing is known about
the variables, which means the operation has to be residualized.

If the argument $u$ of a \texttt{get} operation is mapped to something in the static
heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
operation needs to be emitted.

If the first argument $v$ to a \texttt{set} operation is mapped to something in the
static heap, then the \texttt{set} can performed at optimization time and the static heap
is updated. Otherwise the \texttt{set} operation needs to be emitted. This needs to be
done carefully, because the new value for the field stored in the variable $u$
could itself be static, in which case it needs to be lifted first.

I a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
can be performed at optimization time, which means the operation can be removed
if the types match. If the type check fails statically or if the object is not
in the static heap, the \texttt{guard\_class} is put into the residual trace. This also needs to
lift the variable on which the \texttt{guard\_class} is performed.

Lifting takes a variable that is potentially in the static heap and makes sure
that it is turned into a dynamic variable. This means that operations are
emitted that construct an object that looks like the shape described in the
static heap, and the variable is removed from the static heap.

Lifting a static object needs to recursively lift its fields. Some care needs to
be taken when lifting a static object, because the structures described by the
static heap can be cyclic. To make sure that the same static object is not lifted
twice, the liftfield operation removes it from the static heap \emph{before}
recursively lifting its fields.

@@

SPUR, a tracing JIT for C# seems to be able to remove allocations in a similar
way to the approach described here, as hinted at in the technical report
\cite{XXX}. However, no details for the approach and its implementation are
given.

>>>>

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
whose \texttt{intval} field contains the constant -100.

The following operations on $p_{5}$ and $p_{6}$ could then be
optimized using that knowledge:

@@

So far we have only described what happens when static objects are used in
operations that read and write their fields and in guards. When the static
object is used in any other operation, it cannot stay static. For example, when
a static object is stored in a globally accessible place, the object needs to
actually be allocated, as it might live longer than one iteration of the loop
and because the partial evaluator looses track of it. This means that the static
object needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that actually allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well) This means that instead of the jump,
the following operations are emitted:

@@

The final optimized trace of the example can be seen in
Figure~\ref{fig:step1}.
XXX why does it says ``Figure 4.1'' here but ``Figure 4'' in the label?

@@

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\label{fig:semantics}
\caption{The Operational Semantics of Simplified Traces}
\end{figure*}

@@

Traces are lists of operations. The operations considered here are \texttt{new} (to make
a new object), \texttt{get} (to read a field out of an object), \texttt{set} (to write a field
into an object) and \texttt{guard\_class} (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are initialized to a real
location before being read, otherwise the trace is malformed.

@@

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment $E$ and a heap $H$, which are potentially changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would
be malformed. The environment could not directly map variables to object,
because several variables can contain a pointer to the \emph{same} object. Thus
the "indirection" is needed to express sharing.

@@

The \texttt{new} operation creates a new object $(T,\mathrm{null},\mathrm{null})$ on the
heap under a fresh location $l$ and adds the result variable to the environment,
mapping it to the new location $l$.

The \texttt{get} operation reads a field $F$ out of an object, and adds the result
variable to the environment, mapping it to the read location. The heap is
unchanged.

@@

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. Thus the environment of the
optimizer serves a function similar to that of the environment in the semantics: sharing.

The static heap is a partial function from $V^*$ into the
set of static objects, which are triples of a type and two elements of $V^*$.
A variable $v^*$ is in the domain of the static heap $S$ as long as the
optimizer can fully keep track of the object. The image of $v^*$ is what is
statically known about the object stored in it, \ie its type and its fields. The
fields of objects in the static heap are also elements of $V^*$ (or null, for
short periods of time).

When the optimizer sees a \texttt{new} operation, it optimistically removes it and
assumes that the resulting object can stay static. The optimization for all
further operations is split into two cases. One case is for when the
involved variables are in the static heap, which means that the operation can be
performed at optimization time and removed from the trace. These rules mirror
the execution semantics closely. The other case is for when not enough is known about
the variables, and the operation has to be residualized.

If the argument $v$ of a \texttt{get} operation is mapped to something in the static
heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
operation needs to be residualized.

If the first argument $v$ to a \texttt{set} operation is mapped to something in the
static heap, then the \texttt{set} can performed at optimization time and the static heap
updated. Otherwise the \texttt{set} operation needs to be residualized. This needs to be
done carefully, because the new value for the field, from the variable $u$,
could itself be static, in which case it needs to be lifted first.

If a \texttt{guard\_class} is performed on a variable that is in the static heap, the type check
can be performed at optimization time, which means the operation can be removed
if the types match. If the type check fails statically or if the object is not
in the static heap, the \texttt{guard\_class} is residualized. This also needs to
lift the variable on which the \texttt{guard\_class} is performed.

Lifting takes a variable that is potentially in the static heap and makes sure
that it is turned into a dynamic variable. This means that operations are
emitted that construct an object with the shape described in the
static heap, and the variable is removed from the static heap.

Lifting a static object needs to recursively lift its fields. Some care needs to
be taken when lifting a static object, because the structures described by the
static heap can be cyclic. To make sure that the same static object is not lifted
twice, the \texttt{liftfield} operation removes it from the static heap \emph{before}
recursively lifting its fields.

@@

SPUR, a tracing JIT for C\# seems to be able to remove allocations in a similar
way to the approach described here, as hinted at in the technical report
\cite{XXX}. However, no details for the approach and its implementation are
given.
